Does a present bias influence exploratory choice?

### 2017-01-31 doc/cogsci2017/paper.tex
some comments on a first quick read…

\title{Present bias in exploratory choice}

@@

  Balancing exploration and exploitation is difficult, and under-exploration
  appears to be a particularly pressing problem. We propose that one possible cause of
  under-exploration is present bias, in which immediate rewards (like those
  gained from exploitation) loom larger than future rewards (like those
  gained from exploration). This possible cause of under-exploration is not
  addressed by past lab studies, in which choices generally yield token rewards
  that are converted to money at the end of the experiment, removing the
  inter-temporal aspect of the decision-making process. In this paper, we
  develop an exploratory choice task with immediately-consumed rewards. In
  Experiment 1, we show that people understand the task and respond as expected
  to standard decision variables. In Experiment 2, we introduce a condition in
  which there is a temporal delay between choices and outcomes, and test whether
  this increases exploration by reducing present bias. While we find no effect
  of the delay, we discuss plans for future experiments to test the validity and
  robustness of this result.

@@

@@

Under-exploration is often proposed to be the result of myopia: the
underweighting of temporally distant rewards relative to close ones \citep{March1991, Levinthal1993}. While
choosing actions that are expected to immediately maximize reward is inherently
exploitative, exploration often requires the decision-maker to suffer short-term
costs in search of better future options.

@@

The the second form of myopia is that the decision-maker
is aware of the future but simply doesn't care as much about it.
A wealth have studies have shown that this form of myopia is an important factor
in human decision making, and that people are strongly
biased towards  the present. For example, people will often prefer a larger, later monetary
reward to a smaller, sooner reward when both rewards are in the future, but will
switch their preference when the time until both rewards is reduced so that the
sooner reward is immediate \citep{Kirby1995}. With monetary rewards, the delay
or speed-up required to observe preference reversals is usually several days. With non-monetary rewards,
such as the cessation of an annoying noise \citep{Solnick1980}, watching a video
when bored \citep{Navarick1998}, or drinking soda
when thirsty \citep{Brown2009}, a bias towards immediate rewards has been observed on the scale of
minutes or even seconds.

Present bias has the clear potential to cause people to explore too little, by
increasing the perceived immediate gains of exploiting relative to the delayed
gains of exploring. However, there has been no experimental research
examining how people's bias towards the present affects exploratory choice. In
part, this is by limits in experiment design. In most exploratory
choice experiments, each choice and outcome takes fewer than 10 seconds to
complete, making it unlikely for a large bias to develop between the current
trial and future ones. More importantly, regardless of the length of each trial,
rewards are generally not consumed on a trial-by-trial basis. For example, a participant
might recieve ``points'' on each trial, which are converted to money when the
experiment ends. In this sort of paradigm, there is no reason to expect a bias
towards the present choice's rewards over future rewards, as all rewards are
in fact delivered at the same time.

In the present study, we describe our ongoing efforts to address the lack of research on present bias
and exploratory choice. We develop a novel exploratory choice task using
immediately-consumed rewards and present two experiments. In the first, a pilot
experiment, we vary
the expected payoff of exploration across conditions and test whether this affects
participant choice. The results of this experiment show that participants understand the task and respond
as predicted to well-studied variables. In the second, we introduce a
delayed-outcome condition, which we predicted to increase exploration
if people are present biased. We find no evidence that delaying outcomes affects
the rate of exploration. While this may support the conclusion that exploratory
choice is not susceptible to present bias, we also discuss other possible
explanations of this null result and plans for future experiments.

@@

Thirty-one participants completed the experiment, which was conducted over
Amazon Mechanical Turk (AMT) using the psiTurk framework \citep{Gureckis2015a}.
Participants were paid \$5.00 for their participation, with a performance-based
bonus of up to \$3.00. All participants received the full \$3.00 bonus.

@@

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{figures/machinemappictures.png}
\caption{The machine display seen by participants to allow them to keep track of
  the value of each of the six machines and the next time it would be ready to
  make a choice or produce an outcome. \emph{a}: the display seen by
  participants in Experiment 1 and the immediate condition of Experiment 2.
  \emph{b}: the display seen by participants in teh delayed condition of
  Experiment 2.}
\label{fig:machinemappictures}
\end{figure*}

In the main part of the experiment, participants completed a total of 70 work
periods. For the first 10 work periods, participants simply clicked a button to begin the
slider task. After these initial periods, participants gained access to six
``machines'', each of which had a ``current spinner'' and a ``new spinner'' (Figure~\ref{fig:taskpictures}\emph{c}).
Before each work period, participants clicked the circled machine,
were shown the machine and made a choice about which spinner to run. Based on their choices, there was a
chance the machine would perform the work task for the participant, allowing the
participant to watch their chosen video instead.

@@

The new spinner initially showed a question mark. If a participant chose the
new spinner, then (immediately or after four periods) a new spinner was
created and appeared on the machine. In the high-explore condition, the new spinner's gold wedges could cover
anywhere from 25\% of the spinner to 100\% of the spinner. In the low-explore condition, the new spinner's gold wedges could cover
anywhere from 0\% of the spinner to 75\% of the spinner. The new spinner then
spun and, if the arrow landed on a gold wedge, the machine worked.
Additionally, if the new spinner had a greater gold area than the current
spinner, the new spinner was ``saved'' and the current spinner was
updated to the new spinner.  This meant that
while choosing a new spinner carried risk, it could carry long-term
benefits as the current spinner could be improved from its initial value.
Importantly, the value of choosing a new spinner was higher in the high-spinner
condition, because the range of outcomes was shifted towards higher values.

@@

During the machine choices, participants had access to an ``info'' button at the
bottom of the screen that provided reminders about the dynamics of the task. In
addition, before beginning the full experiment, participants completed two
practice phase. First, they performed several trials of practice using the
machines, with the actual work periods removed. Then, they performed three work
periods practicing the slider task.

@@

The results of the regression are shown in Figure~\ref{fig:exp1results}.
Because we observed few trails for each participant, and the current-spinner
values observed for each participant depend on chance and the participant's
choices, we present individual-level model fits rather than the raw choice data.
We find that at the population level, participants are less likely to choose a
new spinner when the value of the current spinner is higher, with posterior
probability $p>.999$. We also find that participants in the high-explore are more likely to explore a new spinner for a
given current spinner value than those in the low-explore condition, with
posterior probability $p>.99$. We found no interaction between condition and the
effect of current-spinner value.

@@

Forty participants completed the experiment, which was conducted over
AMT using the psiTurk framework \citep{Gureckis2015a}.
Participants were paid \$5.00 for their participation, with a performance-based
bonus of up to \$3.00. All participants received the full \$3.00 bonus.

@@

Experiment 2 followed the design of Experiment 1, with the following changes. In
both conditions, the possible winning proportion of a new spinner ranged fro 0\% to 100\% of
the spinner. In the immediate condition the machine ran immediately after the
choice was made, as in Experiment 1. In the delayed condition, each machine was
presented to the participant four work periods before it was schedules to run,
and the participant made a choice at that time. The machine then had to
``process'' for four work periods, thus delaying
the outcome by over 2 minutes. The participant then clicked the machine again to
return to it and observe its outcome, and either perform the slider task or
watch the video. The machine then cooled of for a single period before being
ready for another choice. The timeline for the delayed condition is shown in
Figure~\ref{fig:machinetimeline}\emph{b}.

@@

While introduces difficulties not found in standard decision-making tasks, we
remain excited about the possibility of creating a paradigm for the use of
immediately-consumable rewards in dynamic decision-making tasks. Such a paradigm
could allow not just a test of whether present bias affects exploratory choice,
but a greater focus on a variety of motivational factors often neglected by
cognitive science.

>>>>

\title{Does a present bias influence exploratory choice?}

@@

  Balancing exploration and exploitation is difficult, and across a wide
  variety of situations under-exploration of new alternatives appears prevalent.
  We propose that one possible cause of
  under-exploration is present bias, whereby immediate rewards (like those
  gained from exploitation) loom larger than future rewards (like those
  gained from exploration). This possible cause of under-exploration is not
  addressed by past studies, in which choices generally yield token rewards
  that are converted to money at the end of the experiment, removing the
  inter-temporal aspect of the decision-making process. In this paper, we
  develop an exploratory choice task with immediately-consumed rewards. In
  Experiment 1, we show that people understand the task and respond as expected
  to standard decision variables. In Experiment 2, we introduce a condition in
  which there is a temporal delay between choices and outcomes, and test whether
  this increases exploration by reducing present bias. While we find no effect
  of the delay, \hl{we discuss plans for future experiments to test the validity and
  robustness of this result.}

@@

\hl{This is good but kind of goes deep into these topics which aren't that central
to the current paper... i.e., learned helplessness per-se.  it can be a little distracting
and adds a lot of citations.  incidently the underexploration of product categories
seems relevant as well}

@@

Under-exploration is often proposed to be the result of myopia: the
underweighting of temporally distant rewards relative to close ones \citep{March1991, Levinthal1993}. 
\hl{does this someone undermine the novelty of your framing/prediction?}
While
choosing actions that are expected to immediately maximize reward is inherently
exploitative, exploration often requires the decision-maker to suffer short-term
costs in search of better future options.

@@

The the second form of myopia is that the decision-maker
is aware of the future but simply doesn't care as much about it.
A wealth of studies have shown that this form of myopia is an important factor
in human decision making, and that people are strongly
biased towards  the present. For example, people will often prefer a larger, later monetary
reward to a smaller, sooner reward when both rewards are in the future, but will
switch their preference when the time until both rewards is reduced so that the
sooner reward is immediate \citep{Kirby1995}. With monetary rewards, the delay
or speed-up required to observe preference reversals is usually several days. With non-monetary rewards,
such as the cessation of an annoying noise \citep{Solnick1980}, watching a video
when bored \citep{Navarick1998}, or drinking soda
when thirsty \citep{Brown2009}, a bias towards immediate rewards has been observed on the scale of
minutes or even seconds.

Present bias has the clear potential to cause people to explore too little, by
increasing the perceived immediate gains of exploiting relative to the delayed
gains of exploring. However, there has been no experimental research
examining how people's bias towards the present affects exploratory choice. In
part, this is due to limits in experiment design. In most exploratory
choice experiments, each choice and outcome takes fewer than 10 seconds to
complete, making it unlikely for a large bias to develop between the current
trial and future ones. More importantly, regardless of the length of each trial,
rewards are generally not consumed on a trial-by-trial basis. For example, a participant
might recieve ``points'' on each trial, which are converted to money when the
experiment ends. In this sort of paradigm, there is no reason to expect a bias
towards the present choice's rewards over future rewards, as all rewards are
in fact delivered at the same time.

In the present study, we describe our ongoing efforts to study the effect of present bias
on exploratory choice. We develop a novel exploratory choice task using
immediately-consumed rewards and present two experiments. In the first, we vary
the expected payoff of exploration across conditions and test whether this affects
participant choice. The results of this experiment show that participants understand the task and respond
as predicted to well-studied variables. In the second, we introduce a
delayed-outcome condition, which we predicted to increase exploration
if people are present biased. However, we find no strong evidence that delaying outcomes affects
the rate of exploration. While this may support the conclusion that exploratory
choice is not susceptible to present bias, we also discuss other possible
explanations of this null result and plans for future experiments.

@@

Thirty-one participants completed the experiment, which was conducted over
Amazon Mechanical Turk (AMT) using the psiTurk framework \citep{Gureckis2015a}.
Participants were paid \$5.00 for their participation, with a performance-based
bonus of up to \$3.00. All participants received the full \$3.00 bonus.
\hl{Maybe mention randomization proceedure to the two conditions?}

@@

\hl{Complex task to follow...  it might make sense to overview it and then
divide into subsections.  like "consuption tasks" "choice tasks", etc...}

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{figures/machinemappictures.png}
\caption{The machine display seen by participants to allow them to keep track of
  the value of each of the six machines and the next time it would be ready to
  make a choice or produce an outcome. \emph{a}: the display seen by
  participants in Experiment 1 and the immediate condition of Experiment 2.
  \emph{b}: the display seen by participants in teh delayed condition of
  Experiment 2. \hl{I wonder if you draw and arrow to show the movement of the machines in the queue?}}
\label{fig:machinemappictures}
\end{figure*}

In the main part of the experiment, participants completed a total of 70 work
periods. For the first 10 work periods, participants simply clicked a button to begin the
slider task. After these initial periods, participants gained access to six
``machines'', each of which had a ``current spinner'' and a ``new spinner'' (Figure~\ref{fig:taskpictures}\emph{c}).
Before each work period, participants clicked the circled machine, \hl{Unclear what circled machine means}
were shown the machine and made a choice about which spinner to run. Based on their choices, there was a
chance the machine would perform the work task for the participant, allowing the
participant to watch their chosen video instead.
\hl{I wonder if you just say more directly that you spin either an existing dial (exploit) or a new
dial (explore).  If the machine came up yellow it would do the "work" for you and you could watch
videos.  If not they you had to do the slider tasks yourself}

@@

The new spinner initially showed a question mark. If a participant chose the
new spinner, then (immediately or after four periods) a new spinner was
created and appeared on the machine. In the high-explore condition, the new spinner's gold wedges could cover
anywhere from 25\% of the spinner to 100\% of the spinner. In the low-explore condition, the new spinner's gold wedges could cover
anywhere from 0\% of the spinner to 75\% of the spinner. The new spinner then
spun and, if the arrow landed on a gold wedge, the machine worked.
Additionally, if the new spinner had a greater gold area than the current
spinner, the new spinner was ``saved'' and the current spinner was
updated to the new spinner.  \emph{This meant that
while choosing a new spinner carried risk, it could carry long-term
benefits as the current spinner could be improved from its initial value.}
Importantly, the value of choosing a new spinner was higher in the high-spinner
condition, because the range of outcomes was shifted towards higher values.

@@

During the machine choices, participants had access to an ``info'' button at the
bottom of the screen that provided reminders about the dynamics of the task. In
addition, before beginning the full experiment, participants completed two
practice phases. First, they performed several trials of practice using the
machines, with the actual work periods removed. Then, they performed three work
periods practicing the slider task.

@@

The results of the regression are shown in Figure~\ref{fig:exp1results}.
Because we observed few trials for each participant, and the current-spinner
values observed for each participant depend on chance and the participant's
choices, we present individual-level model fits rather than the raw choice data.
We find that at the population level, participants are less likely to choose a
new spinner when the value of the current spinner is higher, with posterior
probability $p>.999$. We also find that participants in the high-explore are more likely to explore a new spinner for a
given current spinner value than those in the low-explore condition, with
posterior probability $p>.99$. We found no interaction between condition and the
effect of current-spinner value.

\hl{A little discussion here might help frame things}

@@

Forty participants completed the experiment, which was conducted over
AMT using the psiTurk framework.
Participants were paid \$5.00 for their participation, with a performance-based
bonus of up to \$3.00. All participants received the full \$3.00 bonus.
\hl{Randomization to conditions}

@@

Experiment 2 followed the design of Experiment 1, with the following changes. In
both conditions, the possible winning proportion of a new spinner ranged from 0\% to 100\% of
the spinner. In the immediate condition the machine ran immediately after the
choice was made, as in Experiment 1. In the delayed condition, each machine was
presented to the participant four work periods before it was schedules to run,
and the participant made a choice at that time. The machine then had to
``process'' for four work periods, thus delaying
the outcome by over 2 minutes. The participant then clicked the machine again to
return to it and observe its outcome, and either perform the slider task or
watch the video. The machine then cooled of for a single period before being
ready for another choice. The timeline for the delayed condition is shown in
Figure~\ref{fig:machinetimeline}\emph{b}.

@@

%%% this is a little too apologetic. who cares about our excitement! ;)
%While introduces difficulties not found in standard decision-making tasks, we
%remain excited about the possibility of creating a paradigm for the use of
%immediately-consumable rewards in dynamic decision-making tasks. Such a paradigm
%could allow not just a test of whether present bias affects exploratory choice,
%but a greater focus on a variety of motivational factors often neglected by
%cognitive science.
